---
title: "data-vis"
output: html_document
---

Start by loading the tidyverse packages

```{r setup, message = FALSE, warning = FALSE}
# see https://www.tidyverse.org/packages/
# if not installed yet, run install.packages first
# install.package("gganimate")
# install.package("gifski")
# install.package("png")
# install.package("plotly")
# install.package("highcharter")
# install.package("corrrplot")

library(tidyverse) 
library(timetk)
library(highcharter)
library(readxl)
library(corrplot)
library(scales)
library(plotly)
library(lubridate)
library(janitor)
library(plotly)
library(highcharter)
library(gganimate)
library(gifski)
library(png)
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = NA)
```

### Data visualization with ggplot
    

https://ggplot2.tidyverse.org/

A little background

    + part of the tidyverse and works well with tidy data
    + grammar of graphics
    + most popular data vis package
    + layers and geoms
    + my strong opinion: skip learning base R visualization tools :) 

### Start by loading and wrangling data


```{r}
library(readxl)
url <- "http://www.freddiemac.com/fmac-resources/research/docs/State_and_US_SA.xls"
destfile <- "State_and_US_SA.xls"
curl::curl_download(url, destfile)

states_hpi <- read_excel(destfile, col_types = c("text", 
    "numeric", "numeric", "numeric", "numeric", 
    "numeric", "numeric", "numeric", "numeric", 
    "numeric", "numeric", "numeric", "numeric", 
    "numeric", "numeric", "numeric", "numeric", 
    "numeric", "numeric", "numeric", "numeric", 
    "numeric", "numeric", "numeric", "numeric", 
    "numeric", "numeric", "numeric", "numeric", 
    "numeric", "numeric", "numeric", "numeric", 
    "numeric", "numeric", "numeric", "numeric", 
    "numeric", "numeric", "numeric", "numeric", 
    "numeric", "numeric", "numeric", "numeric", 
    "numeric", "numeric", "numeric", "numeric", 
    "numeric", "numeric", "numeric", "numeric"), 
    skip = 5)

states_hpi_wrangled <- 
  states_hpi %>% 
  # clean up column names from janitor package
  clean_names() %>% 
  slice(1:535) %>% 
  mutate(date = ymd(parse_date_time(month, "%Y%m"))) %>%
  select(date, contains("united"),  everything(), -month) %>% 
  rename(wash_dc = dc, usa = united_states_seasonally_adjusted) 


hpi_tidy <- 
  states_hpi_wrangled %>% 
  gather(state, hpi, -date) %>% 
  group_by(state)

hpi_pct <- 
  hpi_tidy %>% 
  group_by(state) %>% 
  mutate(pct_change = (hpi / lag(hpi)) - 1,
         pct_change_12_mons = (hpi / lag(hpi, 12)) - 1) %>%
  na.omit()

hpi_pct %>% 
  head()
```

### First ggplot!


```{r}
hpi_pct %>% 
  ggplot(aes(x = date, y = pct_change, color = state))
 # ggplot(aes(x = date, y = pct_change, color = state))
```

hmmm, a blank chart, not what we had in mind.


# line chart

```{r}
hpi_pct %>% 
  ggplot(aes(x = date, y = pct_change, color = state)) +
  geom_line()
  
```

# faceting 
```{r}
hpi_pct %>% 
ggplot(aes(x = date, y = pct_change, color = state)) +
  geom_line() +
  # theme(plot.title = element_text(hjust = 0.5)
  #       legend.position = "none",
  #       axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  facet_wrap(~state, ncol = 3) +
  labs(x = "", y = "better label", title = "good title") 
```

# scatter plot

```{r}
hpi_pct %>% 
#  filter(state %in% c("ga", "ny", "fl")) %>% 
ggplot(aes(x = pct_change_12_mons, y = pct_change, color = state)) +
  geom_point() + #alpha = .5
  #facet_wrap(~state, ncol = 10) +
  labs(x = "", y = "better label", title = "scatter") +
  theme(#legend.position = "none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        plot.title = element_text(hjust = 0.5))
```

# scatter with LM

```{r}
hpi_pct %>% 
  filter(state %in% c("ga", "ny", "fl")) %>% 
ggplot(aes(x = pct_change_12_mons, y = pct_change, color = state)) +
  #geom_point() + #alpha = .5
  geom_smooth(method = "lm", se = FALSE, color = "purple", size = 1) +
  geom_point(alpha = .2) +
  facet_wrap(~state, ncol = 3) +
  labs(x = "", y = "better label", title = "scatter with lm") +
  theme(#legend.position = "none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        plot.title = element_text(hjust = 0.5))
```

# Histogram

```{r}
hpi_pct %>% 
  filter(state %in% c("ga", "ny", "fl")) %>% 
  ggplot(aes(x = pct_change)) +
  geom_histogram(color = "blue", fill = "pink", bins = 50) +
  #facet_wrap(~state) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  theme_minimal()
```

# density

```{r}
hpi_pct %>% 
  filter(state %in% c("ga", "ny", "fl")) %>% 
  ggplot(aes(x = pct_change)) +
  geom_density(color = "cornflowerblue") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  theme_minimal() 
```

# density and histogram

```{r}
hpi_pct %>% 
  filter(state %in% c("ga", "ny", "fl")) %>% 
  ggplot(aes(x = pct_change)) +
  geom_histogram(color = "blue", fill = "pink", bins = 50) +
  geom_density(color = "red") +
  facet_wrap(~state) +
  #geom_histogram(color = "blue", fill = "pink", bins = 50) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

# create and save aesthetics

```{r}

my_aesthetics <- 
  theme(plot.title = element_text(colour = "cornflowerblue"),  
        strip.text.x = element_text(size = 8, colour = "white"), 
        strip.background = element_rect(colour = "white", fill = "cornflowerblue"), 
        axis.text.x = element_text(colour = "cornflowerblue"), 
        axis.text = element_text(colour = "cornflowerblue"), 
        axis.ticks.x = element_line(colour = "cornflowerblue"), 
        axis.text.y = element_text(colour = "cornflowerblue"), 
        axis.ticks.y = element_line(colour = "cornflowerblue"),
        axis.title = element_text(colour = "cornflowerblue"),
        legend.title = element_text(colour = "cornflowerblue"),
        legend.text = element_text(colour = "cornflowerblue"))


hpi_pct %>% 
  filter(state %in% c("ga", "ny", "fl")) %>% 
  ggplot(aes(x = pct_change_12_mons, y = pct_change, color = state)) +
  geom_point() + #alpha = .5
  #geom_smooth(method = "lm", se = TRUE, color = "purple", size = 1) +
  facet_wrap(~state, ncol = 10) +
  labs(x = "", y = "better label", title = "scatter with lm") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        plot.title = element_text(hjust = 0.5)) +
  my_aesthetics
```


# Compare two states

```{r}
hpi_pct %>% 
  filter(state == "fl"  | state == "ga" | state == "ca") %>% 
  ggplot(aes(x = date, y = pct_change, color = state)) +
  geom_line() +
  scale_x_date(date_breaks = "3 years", 
               date_labels = "%Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust=0)) +
  # labs()
  ylab("") +
  xlab("") +
  ggtitle("Florida v. GA Housing Price appreciation", 
          subtitle = "more info") +
  labs(caption = "data source: fmac")
```

# Create a recessions shade object

```{r}
# recession plot
recessions <- 
tribble(
  ~Peak, ~Trough,
  "1948-11-01", "1949-10-01",
  "1953-07-01", "1954-05-01",
  "1957-08-01", "1958-04-01",
  "1960-04-01", "1961-02-01",
  "1969-12-01", "1970-11-01",
  "1973-11-01", "1975-03-01",
  "1980-01-01", "1980-07-01",
  "1981-07-01", "1982-11-01",
  "1990-07-01", "1991-03-01",
  "2001-03-01", "2001-11-01",
  "2007-12-01", "2009-06-01"
  ) %>% 
  mutate(Peak = ymd(Peak),
         Trough = ymd(Trough))


recession_shade <- 
  geom_rect(data = recessions, 
            inherit.aes = F, 
            aes(xmin = Peak, 
                xmax = Trough, 
                ymin = -Inf, 
                ymax = +Inf), 
            fill = 'pink', 
            alpha = 0.5)


```

# time series with recessions

```{r, message=FALSE}
hpi_pct %>% 
  filter(state == "fl"  | state == "ga") %>% 
  ggplot(aes(x = date, y = pct_change, color = state)) +
  geom_line() +
  scale_x_date(date_breaks = "2 years", 
               date_labels = "%Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust=0)) +
  ylab("") +
  xlab("") +
  ggtitle("Florida v. GA Housing Price appreciation", 
          subtitle = "more info") +
  labs(caption = "data source: fmac") +
  recession_shade +
  xlim(min(hpi_pct$date), max(hpi_pct$date))
```

# columns custom colors

```{r}
hpi_pct %>% 
  filter((state == "fl"  | state == "ga" ) & date > "2000-01-01") %>% 
  mutate(pos_neg_col = case_when(pct_change < 0 ~ "pink", 
                                 TRUE ~ "cornflowerblue")) %>% 
  ggplot(aes(x = date, y = pct_change, color = pos_neg_col)) +
  geom_col() +
  scale_colour_identity() +
  facet_wrap(~state)
```

# column chart ordered

```{r}
hpi_pct %>% 
  summarise(mean_pct_change = mean(pct_change),
            sd_pct_change = sd(pct_change),
            min_pct_change = min(pct_change)) %>%
  #arrange(desc(mean_pct_change)) %>% 
  top_n(25) %>%  
ggplot(aes(x = reorder(state, mean_pct_change) , y = mean_pct_change, fill = state)) +
  geom_col(width = .5) +
  geom_text(aes(label = state), nudge_y = .0001) +
  theme(legend.position = "none",
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(x = "") +
  scale_y_continuous(labels = scales::percent) 

```

# repel labels

```{r}
library(ggrepel)
hpi_pct %>% 
  summarise(mean_pct_change = mean(pct_change),
            sd_pct_change = sd(pct_change),
            min_pct_change = min(pct_change)) %>%
 arrange(desc(mean_pct_change)) %>% 
  top_n(50) %>% 
ggplot(aes(x = reorder(state, mean_pct_change), y = mean_pct_change, fill = state)) +
  geom_col(width = .5) +
  geom_text_repel(aes(label = state), nudge_y = .0001) +
  theme(legend.position = "none",
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(x = "") +
  scale_y_continuous(labels = scales::percent) 
```


# Plotly

```{r}
library(plotly)
ggplotly(
hpi_pct %>% 
  filter((state == "fl"  | state == "ga" )& date > "2010-01-01") %>% 
  mutate(pos_neg_col = case_when(pct_change < 0 ~ "pink", 
                                 TRUE ~ "cornflowerblue")) %>% 
  ggplot(aes(x = date, y = pct_change, color = pos_neg_col)) +
  geom_col() +
  scale_colour_identity() +
  facet_wrap(~state)
)
```

```{r}
ggplotly(
hpi_pct %>% 
  filter(state == "fl"  | state == "ga") %>% 
  ggplot(aes(x = date, y = pct_change, color = state)) +
  geom_line() +
  scale_x_date(date_breaks = "2 years", 
               date_labels = "%Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust=0)) +
  ylab("") +
  xlab("") +
  ggtitle("Florida v. GA Housing Price appreciation", 
          subtitle = "more info") +
  labs(caption = "data source: fmac") +
  xlim(min(hpi_pct$date), max(hpi_pct$date))
)
```

```{r}
ggplotly(
hpi_pct %>% 
  filter(state %in% c("ga", "ny", "fl")) %>% 
ggplot(aes(x = pct_change_12_mons, y = pct_change, color = state)) +
  geom_point() + #alpha = .5
  geom_smooth(method = "lm", se = TRUE, color = "purple", size = 1) +
  facet_wrap(~state, ncol = 3) +
  labs(x = "", y = "better label", title = "scatter with lm") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
)
```

# Highcharter

Doesn't have a smooth faceter but I like this pacakge and it's widely used in industry.

```{r}
hpi_pct %>% 
  filter(state == "fl"  | state == "ga" | state == "ca") %>% 
  hchart(., hcaes(x = date, y = pct_change, group = state), type = "scatter") %>% 
  hc_title(text = "State HPIs") %>% 
  hc_exporting(enabled = TRUE)
```

Highcharter and plotly are part of a family of R packages called htmlwidgets. They are wrappers for JavaScript libraries - a huge and growing part fo the R ecosystem are these data visualization capabilities. 


# correlation plot

```{r}
library(corrplot)
  hpi_pct %>% 
  select(state, pct_change, date) %>% 
  spread(state, pct_change) %>% 
  select(-date) %>%
  cor() %>% 
  corrplot(method = "circle") # square, color, number, circle, ellipse
```


# gganimate

```{r}
library(gganimate)
library(gifski)
library(png)

hpi_pct %>% 
  filter(state == "fl"  | state == "ga") %>% 
  ggplot(aes(x = date, y = pct_change, color = state)) +
  geom_line() +
  scale_x_date(date_breaks = "2 years", 
               date_labels = "%Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust=0)) +
  ylab("") +
  xlab("") +
  ggtitle("Florida v. GA Housing Price appreciation", 
          subtitle = "more info") +
  labs(caption = "data source: fmac") +
  recession_shade +
  xlim(min(hpi_pct$date), max(hpi_pct$date)) +
  transition_reveal(date)

gganimate::anim_save("hpi-time.gif")
```


```{r}
animate(
hpi_pct %>% 
  summarise(mean_pct_change = mean(pct_change),
            sd_pct_change = sd(pct_change),
            min_pct_change = min(pct_change)) %>%
ggplot(aes(x = state, y = mean_pct_change, fill = state)) +
  geom_col() +
  geom_text_repel(aes(label = state), nudge_y = .0001) +
  theme(legend.position = "none",
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(x = "") +
  scale_y_continuous(labels = scales::percent) +
  transition_states(state, wrap = FALSE) +
  shadow_mark(), 
nframes = 103)
```



```{r}
animate(
hpi_pct %>% 
  filter((state == "fl"  | state == "ga" )& date > "2010-01-01") %>% 
  mutate(pos_neg_col = case_when(pct_change < 0 ~ "pink", 
                                 TRUE ~ "cornflowerblue")) %>% 
  ggplot(aes(x = date, y = pct_change, color = pos_neg_col)) +
  geom_col() +
  scale_colour_identity() +
  facet_wrap(~state) +
  transition_states(date, wrap = FALSE) +
  shadow_mark(), 
nframes = 300)
```
